Certifying, incremental type checking

### 2012-06-12 doc/papers/LFMTP2012/paper.tex
paper

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of program execution --- they ``do not go wrong'', and programming
practices --- modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator:

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly referred as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\vdash M : A \\ A\leq B}{\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task is
highly non-trivial and non-modular: for instance, how to infer the
type of a System \sysname{T} recursor $\recb M N x y P$, in presence
of subtyping? The problem is that we have to guess a type $A$ more
general than that of $N$ and $P$, while giving that type to variable
$\var y$. A well-known algorithm if the subtyping relation forms a
complete lattice is to iterate the typing of $P$ with the join of the
previously computed and the new type until a fixpoint is reached.

@@

Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y} u v
{\fget{\gsubst{\md_4}{\msubst H {\finfer{\var u}}}}}$, $\md_6$ for the
arguments $\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

@@

A few days earlier, the teacher is preparing the tool for the students
to train composing System \sysname{T_{<:}} programs. He (or she)
provides the abstract syntax, the typing rules and the untrusted
typing algorithm in order to build the incremental type checker. How
should he write the algorithm? We know that we want to return a
derivation, and take advantage of a higher-order representation of the
syntax, so it shouldn't be necessary to thread any explicit
environment. The definition should then start with:

@@

\begin{mathleft}
  \quad\caseb{\app{M}N} \\
  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
  \quad\quad
  \match{\md_{\leq}} \\
  \quad\quad\quad\caseb{\infer{ }{\vdash A\leq A}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
  \quad\quad\quad\caseb{\_}
  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
      : A_1}}{\vdash \app M N : B}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\caseb{\tlam x A M} \\
  \quad\quad
  \letdin{\pair B \md}{\env {$\md_{\var x}$} {(\vdash {\var x} : A)}}
  {\finfer {\gsubst M {\msubst x {\fget{\pair A {\md_{\var x}}}}}}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\caseb{\recb M N x y P} \\
  \quad\quad
  \letd {\pair {A_M} {\md_M}} {\finfer{M}} \\
  \quad\quad\letd {\md_{A_M}} {{A_M} \leq \cst{nat}} \\
  \quad\quad
  \letd {\pair {A_N} {\md_N}} {\finfer{N}} \\
  \quad\quad
  \syntax{let} {\pair {A_P} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_N)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A_N}{\md_y}}}}}} \syntax{in}\\
  \quad\quad\letd {\pair A {\pair{\md_{A_N}}{\md_{A_P}}}} {{A_N} \sqcap {A_P}} \\
  \quad\quad
  \syntax{let} {\pair {\_} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A}{\md_y}}}}}} \syntax{in}\\
  \hspace{-2em}\pair {A} {
    \infer{
      \infer*{\md_M \and \md_{A_M}}
      {\vdash M : A} \and
      \infer*{\md_N \and \md_{A_N}}
      {\vdash N : A}\and
      \infer*{\infer*{}{\mbox{$
        \begin{array}{c}
          [\md_x] [\md_y] \\
          \md_P
      \end{array}
        $}} \and \md_{A_P}}{\vdash P : A}
    }{
      \vdash \recb M N x y P : A
    }
  }
\end{mathleft}

@@

The type-checker is now done. There is no case for variables, as we
won't ever meet this case: we already substituted all variables with
their result derivation. Evaluation only need to make sure that in
this case, the derivation will be returned, \ie\ that it reduces the
following way:
$$
\finfer{\fget{\pair A \md}} = \pair A \md
$$
Now $\fget{}$ can be derived automatically from the type of
$\finfer{}$: it is its inverse. It has definition:

% TODO confirmer qu'on met bien la def. des f^0
\begin{mathleft}
  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
    M:A)} \cst{tm} =
  \\\qquad\quad \lamd M \lamd A \lamd \md M
\end{mathleft}

@@

We exploit here an idea similar to run-time \emph{contracts}
\cite{wadler2009well}: there is no static guarantee that computations
won't return ill-typed values, but arguments and results are checked
at run-time at the boundaries of defined functions. This allows to
omit formal justifications, like the fact that we need to iterate
twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
on $\md_x$ in the $\lambda$ case. Of course this incurs on the
performance; we will see a compensation for it by the slicing
mechanism described below in \ref{sec:repr}.

@@

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
% TODO: vérif ce paragraphe après avoir parlé de fcts inv
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would hit the unhandled case $\finfer{\var x}$. This
requires evaluating the program deltas with a carefully designed
strategy (see \ref{sec:comput}).

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. % The requirements are:
% \begin{itemize}
% \item conciseness of the certificates, since they are stored or even
%   transferred remotely;
% \item a well-understood, short and independent verification algorithm.
%   This is known as the \emph{de Bruijn principle};
% \item universality with respect to the particular problem: having a
%   certificate language for each domain of application would just shift
%   the problem of trust; having one universally accepted certificate
%   language that can be tailored to many problems puts a much stronger
%   trust on the unique verifier.
% \end{itemize}
The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is one of the well-accepted solution to this question. It
is a typed lambda-calculus and can be viewed as a language for
\emph{representing} terms, derivations and statements of $\Pi_0^1$
meta-theorems (but not their proofs) in a given inference system
supporting hypothetical judgments \cite{pfenning2001logical}. For
these, it provides built-in support for encoding languages with
binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
\cite{pfenning1988higher}). The user gives a \emph{signature} defining
abstract syntax and inference rules of his \emph{object language}, and
gets an algorithm to verify terms and derivations expressed in the
\LF\ syntax.
% For instance, the rule for System \sysname{T}'s recursor can be
% encoded by the following constant:
% \begin{mathleft}
%   \cst{is\_rec}\ :\
%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
%   \\\quad
%   \prd A {\cst{tp}}
%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
%   \left(
%   \prd x {\cst{tm}} \prd y {\cst{tm}}
%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
%   \app{\cst{is}}\app {(\app P \app x y)} A
%   \right)\to\\\quad
%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
%     P\app x y))} A
% \end{mathleft}
This framework has been used to issue certificates for a number of
practical applications like proof-carrying
code~\cite{necula1997proof}, authentication
framework~\cite{appel1999proof} \ldots %TODO ATP

To write deltas by sharing subterms, we need to be able to address any
subterm. Since a subterm can be open, we must instantiate its free
variables. For this, we define the \emph{slicing} of a well-typed \LF\
term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
Y$, \ldots\ as in \cite{nanevski2008contextual}. Metavariables stand
for well-typed terms defined in a \emph{repository}, and are
instantiated wrt. their free variables. A repository also contains a
distinguished metavariable corresponding to the ``tip'' of the term
stored in it.

We thus introduce Sliced \LF\ (\SLF, see
Fig. \ref{fig:syntax-LF}). Specifically, it is an extension of
\emph{spine canonical \LF} \cite{pfenning2007term} enriched with
metavariables which are references to open terms as in
\cite{nanevski2008contextual}.

\begin{figure}\centering
  \begin{align*}
    K &\gequal \prd x A K \gor \type &
    \text{Kind}\\
    A &\gequal \prd x A A \gor P &
    \text{Type family} \\
    P &\gequal \app {\cst a} S &
    \text{Atomic type} \\
    M &\gequal \lam x M \gor F &
    \text{Canonical object} \\
    F &\gequal \app H S \gor \smeta X \sigma &
    \text{Atomic object} \\
    H &\gequal \var x \gor \cst c &
    \text{Head}\\
    S &\gequal \spinenil \gor \spinecons M S &
    \text{Spine}\\
    \sigma &\gequal \msubstnil \gor \msubstcons \sigma x M &
    \text{Parallel substitution} \\
    \Gamma &\gequal \envnil \gor \envcons \Gamma x A &
    \text{Environment}
  \end{align*}
  \caption{Syntax of \SLF}
  \label{fig:syntax-LF}
\end{figure}

In this variant, only $\eta$-long and $\beta$-normal values have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
\emph{spine} of objects) as opposed to the more standard binary
application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}. The definition of hereditary
substitution $\gsubst M \sigma$ is skipped and can be found in
\cite{pfenning2007term} (it is extended to metavariables as in
\cite{nanevski2008contextual}).

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), as in \eg\ \cite{pierce2000local}:
canonical objects are checked against a given type, whereas we can
synthesize the type of atomic objects; the coercion from the first
grammatical category to the second is where we verify equality between
the synthesized and the checked types.

As is customary, we adopt the Barendregt convention for naming bound
variables, write $A\to A'$ for $\prd x A A'$ when $\var x\notin
\FV{A'}$, $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently
list, map or set notations for $\sigma$ and $\Gamma$. The identity
substitution $\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var
  x}\ |\ \var x \in\dom\Gamma\}$. Metavariables $\smeta X \sigma$
stand for open objects with free variables in $\dom{\sigma}$; To each
use of them is attached a substitution $\sigma$ defining these free
variables. We write $\FMV{M}$ for the set of metavariables in $M$ and
speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.

@@

\begin{definition}
  A \emph{repository} $\mr$ is a finite map from metavariable to
  judgments of atomic object well-typing, along with a distinguished
  metavariable, corresponding to the \emph{tip} of the term contained
  in it:
$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P)), \smeta X
\sigma $$
\end{definition}

The algorithm will maintain the invariant that a repository $\mr$ is
\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint of
$\FMV{\meta X}$ does not contain $\meta X$. If $\mr = (\Delta,
X[\sigma])$, we write $\hat {\mr}$ for the tip $X[\sigma]$, abusively
$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
if $\Delta(\meta X) = (\Gamma\vdash M : A)$.

\begin{definition}
  We define the partial \emph{checkout} operation $\checkout{\mr}$
  (resp. $\checkoutr{M}$, $\checkoutr{S}$) from a repository
  (resp. object, spine) to a metaclosed object as unfolding the
  definitions of all metavariables recursively:
  \begin{align*}
    \checkout{\mr} &= \checkoutr{\hat{\mr}} \\
    \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
    \checkoutr{\lam x M} &= \lam x \checkoutr{M} \\
    \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
    \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} \\
    \checkoutr{\spinenil} &= \spinenil
  \end{align*}
\end{definition}

\begin{example}
  The repository
$$
\small
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
checks out as the metaclosed atomic object
$$
\fbox{$
  \app {\cst{lam}} \lam x
  \fbox{$
    \app {\cst{lam}} \lam y
    \fbox{$
      \app {\cst{app}} \app {\var y}
      (\fbox{$
        \app{\cst{app}}\app{\var x}{\cst o}
      $})
    $}
  $}
$}
$$
where all referable slices have been boxed.
\end{example}

@@

Secondly, we introduce the \CL\ language to transform our \SLF\
objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
other, and deconstruct them by case analysis on an applicative
term. These objects are \SLF\ objects, except that they can contain
special function symbols $\fct f$ and $\fcti f n$. Hence, we name them
with underlined metavariables $\cl F$ instead of $F$.

\begin{figure}
  \centering
  \begin{align*}
    T &\gequal \lam x T \gor
    U & \text{Term} \\
    U &\gequal {\cl F} \gor
    \matchin U \Gamma C & \text{Atomic term}\\
    C &\gequal \casebnobar P U \gor
    C\ \caseb P U & \text{Branches}\\
    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern} \\
    \cl M &\gequal \lam x {\cl M} \gor \cl F &
    \text{Canonical object} \\
    \cl F &\gequal \app {\cl H} {\cl S} \gor \smeta X \sigma &
    \text{Atomic object} \\
    \cl H &\gequal \var x \gor \cst c \gor \fct f \gor \fcti f n&
    \text{Head}\\
    \cl S &\gequal \spinenil \gor \spinecons {\cl M} {\cl S} &
    \text{Spine}\\
    \cl\sigma &\gequal \msubstnil \gor \msubstcons {\cl\sigma} x {\cl M} &
    \text{Parallel substitution} \\
  \end{align*}

@@

  \fbox{$\jh\Delta\Gamma H A L$}
  \qquad
  {Head}

  \begin{mathpar}
    % constant
    \infer[HConst]{
      \Sigma(\cst c) = A; L
    }{
      \jh\Delta\Gamma {\cst c} A L
    }

>>>>

Verifying the well-typing of real-world programs is a critical matter:
on it relies their safety, and it encourages good programming
practices: modularity, early bug catching\ldots. But it is a complex
operation, for the compiler designer and for the human operator:

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for mere ``type checking'', or explicit
``church-style'' calculi. Take for example dependent type systems,
relying on a certain notion of definitional equality on types, or type
systems with subtype polymorphism: both rely on a rule of the form
$$
\infer{\vdash M : A \\ A\leq B}{\vdash M : B}
$$
(for a given notion of $\leq$) which is not syntax-directed. The
implementer has to convert the declarative system to an algorithm, and
convince himself that both are equivalent. This task is highly
non-trivial and non-modular: for instance, how to infer the type of a
System \sysname{T} recursor $\recb M N x y P$, in presence of
subtyping? The problem is that we have to guess a type $A$ more
general than that of $N$ and $P$, while giving that type to variable
$\var y$. A well-known algorithm if the subtyping relation forms a
complete lattice is to iterate the typing of $P$ with the join of the
previously computed and the new type until a fixpoint is reached.

@@

Let us call $\md_5$ the subderivation generated for the definition of
$\var{add}$, $\md_6$ for the arguments
$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

@@

Earlier, the teacher is preparing the tool for the students. He (or
she) provides the abstract syntax, the typing rules and the untrusted
typing algorithm in order to build the incremental type checker. How
should he write the algorithm? We want it to return a derivation, and
take advantage of a higher-order representation of the syntax, so it
shouldn't be necessary to thread any explicit environment. The
definition should then start with:

@@

\begin{mathleft}
  \quad\caseb{\app{M}N} \\
  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}}
  \letd {\pair {A_2} {\md_2}} {\finfer{N}}
  \letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
  \quad\quad
  \match{\md_{\leq}} \\
  \quad\quad\quad\caseb{\infer{ }{\vdash A\leq A}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
  \quad\quad\quad\caseb{\_}
  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
      : A_1}}{\vdash \app M N : B}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\caseb{\tlam x A M}
  \letdin{\pair B \md}{\env {$\md_{\var x}$} {(\vdash {\var x} : A)}}
  {\finfer {\gsubst M {\msubst x {\fget{\pair A {\md_{\var x}}}}}}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\caseb{\recb M N x y P} \\
  \quad\quad
  \letd {\pair {A_M} {\md_M}} {\finfer{M}}
  \letd {\md_{A_M}} {{A_M} \leq \cst{nat}}
  \letd {\pair {A_N} {\md_N}} {\finfer{N}} \\
  \quad\quad
  \syntax{let} {\pair {A_P} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_N)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A_N}{\md_y}}}}}} \syntax{in}\\
  \quad\quad\letd {\pair A {\pair{\md_{A_N}}{\md_{A_P}}}} {{A_N} \sqcap {A_P}} \\
  \quad\quad
  \syntax{let} {\pair {\_} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A}{\md_y}}}}}} \syntax{in}\\
  \quad\quad\pair {A} {
    \infer{
      \infer*{\md_M \and \md_{A_M}}
      {\vdash M : A} \and
      \infer*{\md_N \and \md_{A_N}}
      {\vdash N : A}\and
      \infer*{\infer*{}{\mbox{$
        \begin{array}{c}
          [\md_x] [\md_y] \\
          \md_P
      \end{array}
        $}} \and \md_{A_P}}{\vdash P : A}
    }{
      \vdash \recb M N x y P : A
    }
  }
\end{mathleft}

@@

The type-checker is now done. There is no case for variables, as we
won't ever meet this case: we already substituted all variables with
their result derivation. Evaluation only need to make sure that in
this case, the derivation will be returned, \ie\ that equation
$\finfer{\fget{\pair A \md}} = \pair A \md$ holds.

Note that $\fget{}$ can be derived automatically from the type of
$\finfer{}$: it is its inverse. It has definition:

\begin{mathleft}
  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
    M:A)} \cst{tm} =
  \lamd M \lamd A \lamd \md M
\end{mathleft}

@@

We exploit here an idea similar to run-time \emph{contracts}
\cite{wadler2009well}: there is no static guarantee that computations
won't return ill-typed values, but arguments and results are checked
at run-time at the boundaries of defined functions. This allows to
omit formal justifications, like the fact that we need to iterate
twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
on $\md_x$ in the $\lambda$ case, while detecting errors early. Of
course this incurs on the performance; we will see a compensation for
it by the slicing mechanism described below in \ref{sec:repr}.

@@

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special term construct
$\{x:A\}$ denoting free variables is added to the syntax of terms, and
rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
% TODO: vérif ce paragraphe après avoir parlé de fcts inv
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would hit the unhandled case $\finfer{\var x}$. This
requires evaluating the program deltas with a carefully designed
strategy (see \ref{sec:comput}).

@@

The emission of certificates by untrusted applications first raises
the question of the \emph{data structure} to use for these
certificates. % The requirements are:
% \begin{itemize}
% \item conciseness of the certificates, since they are stored or even
%   transferred remotely;
% \item a well-understood, short and independent verification algorithm.
%   This is known as the \emph{de Bruijn principle};
% \item universality with respect to the particular problem: having a
%   certificate language for each domain of application would just shift
%   the problem of trust; having one universally accepted certificate
%   language that can be tailored to many problems puts a much stronger
%   trust on the unique verifier.
% \end{itemize}
The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
for short, is one of the well-accepted solution to this question. It
is a dependently typed lambda-calculus for \emph{representing} terms,
derivations and statements of $\Pi_0^1$ meta-theorems (but not their
proofs) in a given inference system supporting hypothetical judgments
\cite{pfenning2001logical}. For these, it provides built-in support
for encoding languages with binders and hypothesis (a feature called
$\lambda$-tree syntax or HOAS \cite{pfenning1988higher}). The user
gives a \emph{signature} defining abstract syntax and inference rules
of his \emph{object language}, and gets an algorithm to verify terms
and derivations expressed in the \LF\ syntax.
% For instance, the rule for System \sysname{T}'s recursor can be
% encoded by the following constant:
% \begin{mathleft}
%   \cst{is\_rec}\ :\
%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
%   \\\quad
%   \prd A {\cst{tp}}
%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
%   \left(
%   \prd x {\cst{tm}} \prd y {\cst{tm}}
%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
%   \app{\cst{is}}\app {(\app P \app x y)} A
%   \right)\to\\\quad
%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
%     P\app x y))} A
% \end{mathleft}
This framework has been used to issue certificates for a number of
practical applications like proof-carrying
code~\cite{necula1997proof}, authentication
framework~\cite{appel1999proof} \ldots %TODO ATP

To write deltas by sharing subterms, we need to be able to address any
subterm. Since a subterm can be open, we must instantiate its free
variables. For this, we define the \emph{slicing} of a well-typed \LF\
term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
Y$, \ldots\ as in \cite{nanevski2008contextual}. We thus introduce
Sliced \LF\ (\SLF, see Fig. \ref{fig:syntax-LF}). Specifically, it is
an extension of \emph{spine canonical \LF} \cite{pfenning2007term}
enriched with metavariables which are references to open terms as in
\cite{nanevski2008contextual}.

\begin{figure}\centering
  \begin{align*}
    K &\gequal \prd x A K \gor \type &
    \text{Kind}\\
    A &\gequal \prd x A A \gor P &
    \text{Type family} \\
    P &\gequal \app {\cst a} S &
    \text{Atomic type} \\
    M &\gequal \lam x M \gor F &
    \text{Canonical object} \\
    F &\gequal \app H S \gor \smeta X \sigma &
    \text{Atomic object} \\
    H &\gequal \var x \gor \cst c \gor \fct{f} &
    \text{Head}\\
    S &\gequal \spinenil \gor \spinecons M S &
    \text{Spine}\\
    \sigma &\gequal \msubstnil \gor \msubstcons \sigma x M &
    \text{Parallel substitution} \\
    \Gamma &\gequal \envnil \gor \envcons \Gamma x A &
    \text{Environment}
  \end{align*}
  \caption{Syntax of \SLF}
  \label{fig:syntax-LF}
\end{figure}

In this variant, only $\eta$-long and $\beta$-normal objects have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
(\ie\ to a \emph{spine} of objects) as opposed to the more standard
binary application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}. We differ the discussion of
function symbols $\fct f$ to \ref{sec:comput}. A \emph{value} is an
object containing no function symbols. The definition of hereditary
substitution $\gsubst M \sigma$ is skipped and can be found in
\cite{pfenning2007term} (and extended to metavariables as
\cite{nanevski2008contextual}).

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), as in \eg\ \cite{pierce2000local}: canonical objects
are checked against a given type, whereas we can synthesize the type
of atomic objects; the coercion from the first grammatical category to
the second is where we verify equality between synthesized and checked
types.

As is customary, we adopt the Barendregt convention for naming bound
variables, write $A_1\to A_2$ for $\prd x {A_1} {A_2}$ when $\var x\notin
\FV{A_2}$, $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently
list, map or set notations for $\sigma$ and $\Gamma$. The identity
substitution $\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var
  x}\ |\ \var x \in\dom\Gamma\}$. Metavariables $\smeta X \sigma$
stand for open objects with free variables in $\dom{\sigma}$; To each
use of them is attached a substitution $\sigma$ defining these free
variables. We write $\FMV{M}$ for the set of metavariables in $M$ and
speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.

@@

\begin{definition}
  A \emph{repository} $\mr$ is a finite map from metavariable to
  triplets of environments, atomic objects and atomic types (\ie\
  judgments of well-typing), along with a distinguished metavariable,
  corresponding to the \emph{tip} of the term contained in it:
$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P))\times\smeta X
\sigma $$
\end{definition}

The algorithm in \ref{sec:eval} will maintain the invariant that a repository $\mr$ is
\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint of
$\FMV{\meta X}$ does not contain $\meta X$. If $\mr = (\Delta,
X[\sigma])$, we write $\hat {\mr}$ for the tip $X[\sigma]$, abusively
$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
if $\Delta(\meta X) = (\Gamma\vdash M : A)$.

\begin{definition}
  We define the partial \emph{checkout} operation $\checkout{\mr}$
  (resp. $\checkoutr{M}$, $\checkoutr{S}$) from a repository
  (resp. object, spine) to a metaclosed object as unfolding the
  definitions of all metavariables recursively:
  \begin{align*}
    \checkout{\mr} &= \checkoutr{\hat{\mr}} &
    \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
    \checkoutr{\lam x M} &= \lam x \checkoutr{M} &
    \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
    \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} &
    \checkoutr{\spinenil} &= \spinenil
  \end{align*}
\end{definition}

\begin{example}
  The repository
$
(
\meta X \mapsto
\envnil\vdash
\app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} :
\cst{tm},
\meta Y \mapsto
\env u {\cst{tm}} \vdash
\app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
    {\var u}} v {\var y} } :
\cst{tm}
,
\meta Z \mapsto
\envcons {\env u {\cst{tm}}} v {\cst{tm}} \vdash
\app{\app{\cst{app}}{\var v}}{\var u} :
\cst{tm}
,
\meta T \mapsto
\env x {\cst{tm}} \vdash
\app{\cst{app}}\app{\var x}{\cst{o}} :
\cst{tm}
), \smeta X {}
$
checks out as the metaclosed atomic object
$$
\fbox{$
  \app {\cst{lam}} \lam x
  \fbox{$
    \app {\cst{lam}} \lam y
    \fbox{$
      \app {\cst{app}} \app {\var y}
      (\fbox{$
        \app{\cst{app}}\app{\var x}{\cst o}
      $})
    $}
  $}
$}
$$
where all referable slices have been boxed.
\end{example}

@@

Secondly, we introduce the \CL\ language to transform our \SLF\
objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
others, and deconstruct them by case analysis on an applicative
term. The definition $T$ of a function always begins with a series of
$\lambda$.

\begin{figure}
  \centering
  \begin{align*}
    T &\gequal \lam x T \gor
    U & \text{Term} \\
    U &\gequal F \gor
    \matchin U \Gamma C & \text{Atomic term}\\
    C &\gequal \casebnobar P U \gor
    C\ \caseb P U & \text{Branches}\\
    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern} \\
  \end{align*}

@@
